---
title: "Signature score v.s. metadata"
author: "Masashi Fujita"
date: "11/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(stringr)
library(ggpubr)
library(gt)
```

```{r load, echo=F}
# Load file that store metadata and signature gene expression.
meta.sig <- readRDS("3_ALS.meta.sig.rds")
```

# Methods
Use only Spinal_Cord_Cervical of patients with ALS.
```{r als}
meta.sig.als.spinal <- meta.sig %>%
  filter(Disease == "ALS", `Sample Source` == "Spinal_Cord_Cervical")
```

The number of samples was 117.
```{r count.sample, echo=F}
meta.sig.als.spinal %>% count() %>% gt()
```
All tests were Wilcoxon rank sum test. ns, p > 0.05; \*, p <= 0.05; \*\*, p <= 0.01; \*\*\*, p <= 0.001.

```{r gather, echo=F}
# remove prefix "cl" from the names of clusters.
clusters <- c(
  "cl 1", "cl 2", "cl 3", "cl 4", "cl 5 6", "cl 7", "cl 8",
  "cl 9", "cl 10", "cl 11", "cl 12", "cl 13", "cl 15"
)
levels <- clusters %>% str_remove("cl ") %>% str_replace(" ", "/")
meta.sig.als.spinal.gather <- meta.sig.als.spinal %>%
  gather(key = cluster, value = signature_score, starts_with("cl ")) %>%
  mutate(cluster_wo_cl = cluster %>% str_remove("cl ") %>% str_replace(" ", "/") %>% factor(levels = levels))
```

# Sex
None of microglia signatures had statistically significant differences between
male and female.  
```{r sex, echo=F}
meta.sig.als.spinal.gather %>%
  ggboxplot(x = "cluster_wo_cl",
            y = "signature_score",
            fill = "Sex",
            xlab = "Microglia clusters",
            ylab = "Gene expression signature") +
  stat_compare_means(aes(group = Sex), label = "p.signif")

meta.sig.als.spinal %>% count(Sex) %>% gt()
```

# Ethnicity
There was no significant differences between Hispanic and non-Hispanic.  
```{r ethnicity, echo=F}
meta.sig.als.spinal.gather %>%
  filter(Ethnicity != "Unknown") %>%
  mutate(Ethnicity = ifelse(Ethnicity == "Not Hispanic or Latino, Ashkenazi Jewish", "Not Hispanic or Latino", Ethnicity)) %>%
  ggboxplot(
            x = "cluster_wo_cl",
            y = "signature_score",
            fill = "Ethnicity",
            xlab = "Microglia clusters",
            ylab = "Gene expression signature"
          ) +
  stat_compare_means(aes(group = Ethnicity), label = "p.signif")

meta.sig.als.spinal %>% 
  mutate(Ethnicity = ifelse(Ethnicity == "Not Hispanic or Latino, Ashkenazi Jewish", "Not Hispanic or Latino", Ethnicity)) %>%
  count(Ethnicity) %>%
  gt()
```

# Family History
Signatures of microglia clusters 1, 2, 7, 8, 9, and 10 were enriched in
patients with family history of ALS/FTD. Cluster 13 had an opposite trend.
This result suggests contribution of genetics to microglia fractions. One
concern is that only 26 donors reported presence/absence of family history.  
```{r familiy, echo=F}

# THIS CODE CHUNK DOES NOT WORK ANYMORE PROBABLY BECAUSE OF VERSION ISSUE.

# meta.sig.als.spinal.gather %>%
#   filter(`Family History of ALS/FTD?` != "Unknown") %>%
#   ggboxplot(
#             x = "cluster_wo_cl",
#             y = "signature_score",
#             fill = "Family History of ALS/FTD?",
#             xlab = "Microglia clusters",
#             ylab = "Gene expression signature"
#           ) +
#   stat_compare_means(aes(group = `Family History of ALS/FTD?`), label = "p.signif")
# 
# meta.sig.als.spinal %>% 
#   count(`Family History of ALS/FTD?`) %>%
#   gt()
```

# FTD
No significant differences.  
```{r ftd, echo=F}
meta.sig.als.spinal.gather %>%
  filter(`MND with FTD?` %in% c("Yes", "No")) %>%
  ggboxplot(
            x = "cluster_wo_cl",
            y = "signature_score",
            fill = "MND with FTD?",
            xlab = "Microglia clusters",
            ylab = "Gene expression signature"
          ) +
  stat_compare_means(aes(group = `MND with FTD?`), label = "p.signif")

meta.sig.als.spinal %>% 
  count(`MND with FTD?`) %>%
  gt()
```

# Dementia
No significant differences.  
```{r dementia, echo=F}
meta.sig.als.spinal.gather %>%
  filter(`MND with Dementia?` %in% c("Yes", "No")) %>%
  ggboxplot(
            x = "cluster_wo_cl",
            y = "signature_score",
            fill = "MND with Dementia?",
            xlab = "Microglia clusters",
            ylab = "Gene expression signature"
          ) +
  stat_compare_means(aes(group = `MND with Dementia?`), label = "p.signif")

meta.sig.als.spinal %>% 
  count(`MND with Dementia?`) %>%
  gt()
```

# Age at Symptom Onset
No siginificant differences. Dichotomized with the median age at symptom onset
(63 yrs).  
```{r age.onset, echo=F}
meta.sig.als.spinal.gather %>%
  filter(!is.na(`Age at Symptom Onset`)) %>%
  mutate(`Age at Symptom Onset` = ifelse(`Age at Symptom Onset` < 63, "<63 yrs", ">=63 yrs")) %>%
  ggboxplot(
            x = "cluster_wo_cl",
            y = "signature_score",
            fill = "Age at Symptom Onset",
            xlab = "Microglia clusters",
            ylab = "Gene expression signature"
          ) +
  stat_compare_means(aes(group = `Age at Symptom Onset`), label = "p.signif")

cat("Age at Symptom Onset:\n")
print(summary(meta.sig.als.spinal$`Age at Symptom Onset`))
```

# Age at Death
Cluster 11 had a difference. Dichotomized with the median age at death
(66 yrs).  
```{r age.death, echo=F}
meta.sig.als.spinal.gather %>%
  filter(!is.na(`Age at Death`)) %>%
  mutate(`Age at Death` = ifelse(`Age at Death` < 66, "<66 yrs", ">=66 yrs")) %>%
  ggboxplot(
            x = "cluster_wo_cl",
            y = "signature_score",
            fill = "Age at Death",
            xlab = "Microglia clusters",
            ylab = "Gene expression signature"
          ) +
  stat_compare_means(aes(group = `Age at Death`), label = "p.signif")

cat("Age at Death:\n")
print(summary(meta.sig.als.spinal$`Age at Death`))
```

# Disease Duration
```{r duration, echo=F}
meta.sig.als.spinal.gather %>%
  filter(!is.na(`Disease Duration in Months`)) %>%
  ggplot(aes(
    x = `Disease Duration in Months`,
    y = signature_score,
  )) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~ cluster, scales = "free_y") +
  theme_classic() +
  theme(strip.background = element_rect(fill="gray")) +
  xlab("Disease Duration (months)") +
  ylab("Gene expression signature")

calc.cor <- function(t){
  ct <- cor.test(~ signature_score + `Disease Duration in Months`, data=t)
  data.frame(pearson.correlation.coeff = ct$estimate, p.value=ct$p.value)
}

cor.res <- meta.sig.als.spinal.gather %>% 
  group_by(cluster) %>%
  do(calc.cor(.)) %>%
  ungroup

gt(cor.res)

cat("Disease Duration in Months:\n")
print(summary(meta.sig.als.spinal$`Disease Duration in Months`))

```

# Disease Duration (5-year threshold)
A threshold of 5 years seems often used in MS studies. Donors with longer
disease duration have lower signatures of clusters 1, 4, 5/6, and 7.  
```{r duration.5yr, echo=F}
meta.sig.als.spinal.gather %>%
  filter(!is.na(`Disease Duration in Months`)) %>%
  mutate(`Disease Duration` = ifelse(`Disease Duration in Months` >= 60, ">=5 yrs", "<5 yrs")) %>%
  ggboxplot(
            x = "cluster_wo_cl",
            y = "signature_score",
            fill = "Disease Duration",
            xlab = "Microglia clusters",
            ylab = "Gene expression signature"
          ) +
  stat_compare_means(aes(group = `Disease Duration`), label = "p.signif")

cat("Disease Duration in Months:\n")
```

# Disease Duration (2-year threshold)
This threshold (2 years) is shown here because it had more associations with
signatures than other thresholds (e.g. 1 year, 3 years, and 5 years). Donors
with longer disease duration have lower signatures of clusters 1, 4, 7,
8, and 10. Cluster 13 had an opposite trend.  
```{r duration.2yr, echo=F}
meta.sig.als.spinal.gather %>%
  filter(!is.na(`Disease Duration in Months`)) %>%
  mutate(`Disease Duration` = ifelse(`Disease Duration in Months` >= 24, ">=2 yrs", "<2 yrs")) %>%
  ggboxplot(
            x = "cluster_wo_cl",
            y = "signature_score",
            fill = "Disease Duration",
            xlab = "Microglia clusters",
            ylab = "Gene expression signature"
          ) +
  stat_compare_means(aes(group = `Disease Duration`), label = "p.signif")
```

# C9orf72
Cluster 11 had a difference.  
```{r C9orf72, echo=F}

# THIS CODE CHUNK DOES NOT WORK ANYMORE PROBABLY BECAUSE OF VERSION ISSUE.

# meta.sig.als.spinal.gather %>%
#   filter(`C9orf72 Repeat Expansion (Data from CUMC)` %in% c("Yes", "No")) %>%
#   ggboxplot(
#             x = "cluster_wo_cl",
#             y = "signature_score",
#             fill = "C9orf72 Repeat Expansion (Data from CUMC)",
#             xlab = "Microglia clusters",
#             ylab = "Gene expression signature"
#           ) +
#   stat_compare_means(aes(group = `C9orf72 Repeat Expansion (Data from CUMC)`), label = "p.signif")
# 
# meta.sig.als.spinal %>% 
#   count(`C9orf72 Repeat Expansion (Data from CUMC)`) %>%
#   gt()
```

# ATXN2
No significant differences.  
```{r ATXN2, echo=F}

# THIS CODE CHUNK DOES NOT WORK ANYMORE PROBABLY BECAUSE OF VERSION ISSUE.

# meta.sig.als.spinal.gather %>%
#   filter(`ATXN2 Repeat Expansion (Data from CUMC)` != "Unknown") %>%
#   ggboxplot(
#             x = "cluster_wo_cl",
#             y = "signature_score",
#             fill = "ATXN2 Repeat Expansion (Data from CUMC)",
#             xlab = "Microglia clusters",
#             ylab = "Gene expression signature"
#           ) +
#   stat_compare_means(aes(group = `ATXN2 Repeat Expansion (Data from CUMC)`), label = "p.signif")
# 
# meta.sig.als.spinal %>% 
#   count(`ATXN2 Repeat Expansion (Data from CUMC)`) %>%
#   gt()
```
